\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2022-04-13 sections/methods.tex
Change: Update the consensus network section of the Results

  The first method, hereby referred to as ``seqtime'', used generalized Lotka-Volterra (gLV) equations to model the microbial community dynamics and utilized the Klemm–Eguı́luz algorithm to generate a clique-based interaction network~\cite{Rottjers2018}.
  We used the \href{https://github.com/hallucigenia-sparsa/seqtime}{seqtime} R package to simulate communities with number of species ($N$) varying from 10 to 150 (10, 25, 50, 100, 150 and 200).
  The initial species concentrations are randomly sampled from a Poisson distribution and the simulation is rerun to generate a number of samples ($S$) varying from 50 to 500 (50, 100, 200, 500) for different communities.
  The abundance values of the species in the community at the end of the simulation time are used to create the OTU table.

  The second method, hereby referred to as ``NorTA'', used the Normal to Anything (NorTA) approach coupled with a given interaction network topology to generate the abundance distribution of the microbial community~\cite{Kurtz2015}.
  We used the \href{https://github.com/zdk123/SpiecEasi}{spieceasi} R package to simulate communities with different network topologies (scale-free, cluster, block, Erdos-Renyi, band and hub) and target abundance distributions (Negative Binomial, Poisson, Zero-Inflated Negative Binomial).
  The OTU table was generated using the American Gut Project example in the \texttt{spieceasi} package (amgut1.filt) with the default parameter options.

>>>>

  % TODO: Add reference: Karoline Faust, Franziska Bauchinger, Béatrice Laroche, Sophie de Buyl, Leo Lahti, Alex D Washburne, Didier Gonze, Stefanie Widder, "Signatures of ecological processes in microbial community time series", Microbiome (2018).
  The first method, hereby referred to as ``seqtime'', used generalized Lotka-Volterra (gLV) equations to model the microbial community dynamics and utilized the Klemm–Eguı́luz algorithm to generate a clique-based interaction network~\cite{Rottjers2018}.
  We used the \href{https://github.com/hallucigenia-sparsa/seqtime}{seqtime} R package to simulate communities with number of species ($N$) varying from 10 to 150 (10, 25, 50, 100, 150 and 200).
  The initial species concentrations are randomly sampled from a Poisson distribution and the simulation is rerun to generate a number of samples ($S$) varying from 50 to 500 (50, 100, 200, 500) for different communities.
  The abundance values of the species in the community at the end of the simulation time are used to create the OTU table.

  The second method, hereby referred to as ``NorTA'', used the Normal to Anything (NorTA) approach coupled with a given interaction network topology to generate the abundance distribution of the microbial community~\cite{Kurtz2015}.
  We used the \href{https://github.com/zdk123/SpiecEasi}{spieceasi} R package~\cite{Kurtz2015} to simulate communities with different network topologies (scale-free, cluster, block, Erdos-Renyi, band and hub) and target abundance distributions (Negative Binomial, Poisson, Zero-Inflated Negative Binomial).
  The OTU table was generated using the American Gut Project example in the \texttt{spieceasi} package (amgut1.filt) with the default parameter options.

### 2022-04-13 sections/results.tex
Change: Update the consensus network section of the Results

  % TODO: Update after getting SparCC results
  In order to obtain a quantitative picture of the differences between the inferred networks, we analyzed the distribution of common nodes and edges (Figure~\ref{fig:figure5}B) using UpSet plots~\cite{Lex}.
  Only \ac{harmonies}, \ac{cozine},\ac{spieceasi}, \ac{spring}, propr, and \ac{sparcc} are used in the comparison since Pearson and Spearman add a large number of spurious edges since they are not intended for compositional datasets).
  The results for the node intersections show that the networks have a large number of nodes in common ($33$ out of $68$ total nodes) and no network possesses any unique node.
  The edge intersections in contrast show that only $8$ edges (out of $202$ total edges) are in common between all the methods and each network has a large number of unique edges.
  These results indicate that there is a substantial rewiring of connections in the inferred networks.

@@

  \subsubsection*{The consensus network}

  Therefore, we propose the construction of a consensus network (Figure~\ref{fig:figure5}C) involving \ac{mldm}, \ac{spieceasi} and \ac{sparcc}. This consensus network is built by merging the p-values generated from bootstraps of the original taxonomy counts table using the Browns p-value combining method~\cite{brown_400_1975} (see Methods section).

  % FIXME: We have synthetic data
  Similar to the previous steps of the pipeline, where were we evaluated the performance of methods on mock datasets, we can evaluate the network inference algorithms using synthetic interaction data.
  Specifically, we use the ...

  Note that we do not use Pearson and Spearman methods in the p-value merging step and these algorithms are only used for demonstration and comparison.
  Based on this approach, \ac{micone} reports as default output the consensus network, annotated with weights (correlations for \ac{sparcc} and direct associations for the other methods) for all four methods.

@@

  \subsubsection*{The default pipeline}
  % TODO: Change the order of Figure 6 and 7

>>>>

  % TODO: Update after getting SparCC results
  In order to obtain a quantitative picture of the differences between the inferred networks, we analyzed the distribution of common nodes and edges (Figure~\ref{fig:figure5} B and C) using UpSet plots~\cite{Lex}.
  Only \ac{harmonies}, \ac{cozine},\ac{spieceasi}, \ac{spring}, propr, and \ac{sparcc} are used in the comparison since Pearson and Spearman add a large number of spurious edges since they are not intended for compositional datasets).
  The results for the node intersections show that the networks have a large number of nodes in common ($33$ out of $68$ total nodes) and no network possesses any unique node.
  The edge intersections in contrast show that only $8$ edges (out of $202$ total edges) are in common between all the methods and each network has a large number of unique edges.
  These results indicate that there is a substantial rewiring of connections in the inferred networks.

@@

  \subsection*{The consensus network}

  We observe a large number of rewirings in the inferred association networks generated by the different methods (Figure \ref{fig:figure5}C), implying that there is a lack of agreement on the associations that should be a part of the final network.
  Therefore, we have developed a consensus method that would take into consideration the evidence offered by each network inference algorithm and generate a consensus network that would contain the most probable interactions for the input data.
  % TODO: Should we include Pearson and Spearman in the consensus (currently we do)!
  We have developed two algorithms for the generation of the consensus network, the simple voting and the scaled-sum methods (see Methods).
  First, for the correlation-based methods we merge the p-values using the Brown's p-value merging method~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016,faustCoNetAppInference2016} (see Methods).
  Second, we filter all the inferred networks based on an association strength threshold of 0.1 and a p-value cutoff of 0.05.
  Finally, we apply the consensus algorithms we have developed on these filtered networks.
  The simple voting method does a consensus based on whether the number of inferred networks that have an edge between a pair of nodes is greater than a threshold.
  The scaled-sum method does a consensus based on whether the normalized sum of an edge between a pair of nodes from all inferred networks is greater than a threshold.
  The default algorithm for the pipeline is the scaled-sum method, since this method takes into account the weights of the association in the determination of the final consensus.
  Additionally, by default we include all the network inference methods in the consensus calculation by default, but a subset of methods can be considered if it is required.
  Based on this approach, \ac{micone} reports as default output the consensus network, where is each edge is annotated with weights (correlations for the correlation-based methods and direct associations for the other methods) from all the methods used in the consensus algorithm.

  Similar to the previous steps of the pipeline, where were we evaluated the performance of methods on mock datasets, we can evaluate the network inference algorithms and the final consensus network using synthetic interaction data.
  For this purpose, we generated synthetic interaction data using the ``NorTA''[ref] and ``seqtime''[ref] methods (see Methods), which have been previously used for benchmarking network inference methods.
  For each method, an \ac{otu} counts table is generated based on the selected interaction network topology and properties.
  This counts table is used as the input to the \ac{micone} pipeline to generate predicted associations.
  The interaction network used to generate the counts table is used as the source of true interactions to calculate the precision and sensitivity for each network inference algorithm for the NorTA datasets (Figure \ref{fig:figure6}).
  Each point in the scatter plot represents a network with different input network topology and OTU abundance distributions.
  Overall, we see that the scaled-sum method for a parameter value of 0.5 performs the best in terms of average precision ($0.985$).
  Among the individual inference methods, we observe that \ac{spieceasi} has the best average precision ($0.944$)
  The goal of the consensus network algorithms is to increase the precision of the resultant consensus network by removing edges that are either not present in most of the inference methods or whose association strength is low.
  From Figure \ref{fig:figure6} we can confirm that the consensus algorithm, especially the scaled-sum method, does indeed produce networks with more precision, but as a consequence the sensitivity of the networks is reduced.
  Therefore, the consensus networks provide the means to obtain a short list of associations that would have a high probability of being present in the real interaction network.
  This small set of associations could be further tested and validated experimentally to verify their plausibility.

  % TODO: Supplementary work: Figure S6
  The precision and sensitivity graphs on the seqtime data can be found in the Supplementary section.

@@

  \subsection*{The default pipeline}

### 2022-04-13 sections/tables_figures.tex
Add: TODO about figure 3 captions
Change: Update the pipeline section of Results
Change: Update the Network Variability section of Results
Change: Update the consensus network section of the Results

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.67\linewidth]{figure1.pdf}
    \caption{
      \textbf{The workflow of the microbial co-occurrence analysis pipeline}.
      The steps can be grouped into four major groups: \textbf{(DC)} \textbf{D}enoising and \textbf{C}lustering, \textbf{(TA)} \textbf{T}axonomy \textbf{A}ssignment, \textbf{(OP)} \textbf{O}TU or ESV \textbf{P}rocessing, and \textbf{(NI)} \textbf{N}etwork \textbf{I}nference.
      Each step incorporates several processes, each of which in turn have several alternate algorithms for the same task (indicated by the text to the right of the blue boxes).
      The text along the arrows describes the data that is being passed from one step to another. For details on each process and data types, see Methods.
    }
    \label{fig:figure1}
  \end{figure}
  \FloatBarrier
  \newpage

  \begin{figure}[H]
    \centering
    \includegraphics[width=0.85\linewidth]{figure2.pdf}
  \end{figure}
  \begin{figure}[H]
    \centering
      \caption{
      \textbf{The choice of database contributes to the most variance in the networks}.
      \textbf{(A)} The total relative variance in the networks contributed by the DC, TA and OP steps of the pipeline (right) and the linear model used to calculate the relative variance (left), see the Methods section for details.
      \textbf{(B)} All combinations of inferred networks are shown as points on a PCA plot.
      The color of the points corresponds to the taxonomy database, the shape corresponds to the denoising/clustering method and the size corresponds to whether low abundance OTUs were removed or not.
      \textbf{(B inset)} The network generated using DC=dada2, TA=GG, OP=no and NI=SPARCC and represents the particular point shown (big red square).
      The points on the PCA plot can be separated based on the TA step and the differences due to the DC and OP steps are comparatively not as significant.
    }
    \label{fig:figure2}
  \end{figure}
  \FloatBarrier
  \newpage

  \begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figure3.pdf}
    \caption{
      \textbf{The representative sequences generated by the different denoising/clustering methods are very similar but differ in the sequences that are in low abundance.}
      \textbf{(A)} The average weighted UniFrac distance between the representative sequences shows that the representative sequences and their compositions are fairly identical between the methods,
      \textbf{(B)} The relatively larger average unweighted UniFrac distance indicates that methods differ in their identification of sequences of low abundance,
      \textbf{(C, D)} The distributions of the average weighted UniFrac distance between the expected sequence profile and the calculated sequence profile in mock datasets.
      \textbf{(D)} The distributions of the average unweighted UniFrac distance show that dada2 and Deblur were the best performing methods in most of the datasets.
    }
    \label{fig:figure3}
  \end{figure}
  \FloatBarrier
  \newpage

@@

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{figure6.pdf}
    \caption{
      \textbf{Network inference and taxonomic assignment have the highest influence on the inferred network structures.}
      \textbf{(A)} The network constructed using the default pipeline parameters (DC=\ac{dada2}, TA=\ac{ncbi}, OP=on, NI=\ac{sparcc}) is compared with networks generated when one of the steps use a different tool.
      The common connections (common with the default network) are in black, connections unique to the network are colored purple and connections in the default network but not present in the current network are gray.
      \textbf{(B)} The L1 distance between the networks generated by changing one step of the default pipeline and the network generated using the default parameters.
    }
    \label{fig:figure6}
  \end{figure}
  \FloatBarrier
  \newpage

  \begin{figure}[h]
    \centering
    \includegraphics[width=17cm]{figure7.pdf}
    \caption{
      The consensus networks generated using the default pipeline settings.
      \textbf{(A)} Co-occurrence network of the Hard Palate microbiome generated from samples of healthy subjects in a periodontal diseases study.
      \textbf{(B)} Co-occurrence network of the Stool microbiome generated from samples of healthy subjects in a fecal microbiome transplant study.
  }
    \label{fig:figure7}
  \end{figure}

>>>>

  \begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{figure1.pdf}
  \end{figure}
  \begin{figure}[H]
    \centering
    \caption{
      \textbf{The workflow of the microbial co-occurrence analysis pipeline}.
      The steps can be grouped into four major groups: \textbf{(DC)} \textbf{D}enoising and \textbf{C}lustering, \textbf{(TA)} \textbf{T}axonomy \textbf{A}ssignment, \textbf{(OP)} \textbf{O}TU or ESV \textbf{P}rocessing, and \textbf{(NI)} \textbf{N}etwork \textbf{I}nference.
      Each step incorporates several processes, each of which in turn have several alternate algorithms for the same task (indicated by the text to the right of the blue boxes).
      The text along the arrows describes the data that is being passed from one step to another. For details on each process and data types, see Methods.
    }
    \label{fig:figure1}
  \end{figure}

  \FloatBarrier
  \newpage

  \begin{figure}[H]
    \centering
    \includegraphics[width=1.0\linewidth]{figure2.pdf}
  \end{figure}
  \begin{figure}[H]
    \centering
      \caption{
      \textbf{The choice of database contributes to the most variance in the networks}.
      \textbf{(A)} The total relative variance in the networks contributed by the DC, TA and OP steps of the pipeline (right) and the linear model used to calculate the relative variance (left), see the Methods section for details.
      \textbf{(B)} All combinations of inferred networks are shown as points on a PCA plot.
      The color of the points corresponds to the taxonomy database, the shape corresponds to the denoising/clustering method and the size corresponds to whether low abundance OTUs were removed or not.
      \textbf{(B inset)} The network generated using DC=dada2, TA=GG, OP=no and NI=SPARCC and represents the particular point shown (big red square).
      The points on the PCA plot can be separated based on the TA step and the differences due to the DC and OP steps are comparatively not as significant.
    }
    \label{fig:figure2}
  \end{figure}
  \FloatBarrier
  \newpage

  % TODO: Describe the datasets being shown here clearly. In the caption + figure legends (if necessary) and in the text
  \begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figure3.pdf}
    \caption{
      \textbf{The representative sequences generated by the different denoising/clustering methods are very similar but differ in the sequences that are in low abundance.}
      \textbf{(A)} The average weighted UniFrac distance between the representative sequences shows that the representative sequences and their compositions are fairly identical between the methods,
      \textbf{(B)} The relatively larger average unweighted UniFrac distance indicates that methods differ in their identification of sequences of low abundance,
      \textbf{(C, D)} The distributions of the average weighted UniFrac distance between the expected sequence profile and the calculated sequence profile in mock datasets.
      \textbf{(D)} The distributions of the average unweighted UniFrac distance show that dada2 and Deblur were the best performing methods in most of the datasets.
    }
    \label{fig:figure3}
  \end{figure}
  \FloatBarrier
  \newpage

@@

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{figure6.pdf}
    \caption{
      The precision and sensitivity of the inferred networks on the synthetic data.
    }
    \label{fig:figure6}
  \end{figure}
  \FloatBarrier
  \newpage

  \begin{figure}[h]
    \centering
    \includegraphics[width=17cm]{figure7.pdf}
    \caption{
      \textbf{Network inference and taxonomic assignment have the highest influence on the inferred network structures.}
      \textbf{(A)} The network constructed using the default pipeline parameters (DC=\ac{dada2}, TA=\ac{ncbi}, OP=on, NI=\ac{sparcc}) is compared with networks generated when one of the steps use a different tool.
      The common connections (common with the default network) are in black, connections unique to the network are colored purple and connections in the default network but not present in the current network are gray.
      \textbf{(B)} The L1 distance between the networks generated by changing one step of the default pipeline and the network generated using the default parameters.
    }
    \label{fig:figure7}
  \end{figure}
